---
title: "How far to the closest hospital?"
author: "Martin Geissmann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: md_document
---

# Intro

Imagine you just fell out of a plane over Switzerland and landed at a random location. As you are wounded your number one concern it to get to a hospital asap. Not any hospital but one that has an emergency room and is equiped at least with a CT (computer tomography).

In the following we want to create a map that shows the *driving time* to the closest emergency hospital. We use the following:

* a public list of Swiss hospitals
* our own OSRM server (Docker) to estimate driving times
* 2D interpolation and mapping functions in R (ggmap, raster, sp etc.)
* leaflet to create an interactive map

# Preparation

## OSRM server

> The Open Source Routing Machine or OSRM is a C++ implementation of a high-performance routing engine for shortest paths in road networks. Licensed under the permissive 2-clause BSD license, OSRM is a free network service. OSRM supports Linux, FreeBSD, Windows, and Mac OS X platform.[^wiki]

There is a [OSRM package for R](https://github.com/rCarto/osrm) that makes it easy to send requests to a OSRM server. It can be installed with the following.

```{r installosrm, eval=F}

# To install the version from CRAN
install.packages("osrm")

```

```{r loadosrm}

library(osrm)
??osrm

```

By default the OSRM package uses a public demo server.

```{r verifyserver}

getOption("osrm.server")

```

The public server is fine for testing but not for sending too many requests in a short time. It will give errors and not work. Therefore it is suggested to get your own instance of OSRM.

The easiest solution is to run it with Docker. This requires 2 preparatory steps:

1. Install Docker CE https://docs.docker.com/install/
2. Get the *osrm-backend* running https://hub.docker.com/r/osrm/osrm-backend/

The first step of installing Docker might take 30 minutes or so. The second step is also pretty straightforward if you follow the *Quick Start* section in the Docker Hub link provided above.

The OpenStreetMap data required to compute paths can be downloaded from http://download.geofabrik.de. For this project we want the latest data (*.osm.pbf) of Switzerland, which can be downloaded (here)[http://download.geofabrik.de/europe/switzerland-latest.osm.pbf].

```{bash wget, eval = F}

wget http://download.geofabrik.de/europe/switzerland-latest.osm.pbf

```

Then run the following:

```{bash rundocker, eval = F}

docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-extract -p /opt/car.lua /data/switzerland-latest.osm.pbf

docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-partition /data/switzerland-latest.osrm
docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-customize /data/switzerland-latest.osrm

docker run -t -i -p 5000:5000 -v "${PWD}:/data" osrm/osrm-backend osrm-routed --algorithm mld /data/switzerland-latest.osrm

```

The last command will run the docker container. You can check it is running with `bash docker container ls`. Also let's request a route from Zurich (47.3769 N, 8.5417 E) to Geneva (46.2044 N, 6.1432 E).

```{bash testit}

curl "http://127.0.0.1:5000/route/v1/driving/8.5417,47.3769;6.1432,46.2044?steps=false"

```

If the output is something with "Bahnhofsplatz" and "Place de Bel-Air" then it is working fine.

Now back in R you want to set the server to your own, so all requests are handled by your own instance. This has to be done each time after you load the `osrm` package.

```{r setserver}

options(osrm.server = "http://localhost:5000/")

# check if it's set correctly
getOption("osrm.server")

```

[^wiki]: https://en.wikipedia.org/wiki/Open_Source_Routing_Machine